Effectful Semantics for Substructural SSA

### 2023-11-15 pldi24/paper.tex
finished first draft of related work

There has been much less work on denotational semantics for SSA, or
directly on its equational theory. \citet{pop-ssa-inout-2009} give an
unusual denotational model of SSA in terms of the iteration structure
of a program, which they use to better understand the loop-closing
$\phi$-nodes found both in the gated SSA representation as well as
practical compilers such as GCC.  \citet{garbuzov-structural-cfg-2018}
exhibit a correspondence between an operational semantics for SSA, and
an operational semantics for the call-by-push-value calculus of
\citet{cbpv}. Despite working over an operational semantics, they use
the normal-form bisimulations of \citet{lassen-bisim} to derive an
equational theory for use in justifying optimisations. \todo{Their
operational semantics does contain not contain any effects, which
means that it cannot justify effect-dependent optimisations such as
eliminating redundant writes.}

We are the first paper to connect a weak memory semantics to the
semantics of SSA. In order to establish a rich equational theory,
we needed a number of features of the semantics which are sometimes
difficult to come by in weak memory semantics. First, sequencing
must be associative (i.e., $(t_1; t_2); t_3 \equiv t_1; (t_2; t_3)$).
Second, the semantics must support loops, including the equation
that a while-loop is equal to its unrolling. Finally, booleans need
to be distributive (i.e., $(A \otimes \mb{2}) \simeq (A + A)$).

\TODO{\citet{leaky-semicolon} for weak memory stuff: future work: add while loops, and also refinement}
\TODO{\citet{paviotti-modular-relaxed-dep-20}}
\TODO{\citet{sparky}}

\todo{Main thing here to say is that we need semantics for while loops, plus associativity of semicolon, if statements must support distributivity}

% \TODO{\cite{schneider-phd}}
% \TODO{\cite{rvsdg}}

\TODO{this}

>>>>

There has been much less work on denotational semantics for SSA, or
directly on its equational theory. \citet{pop-ssa-inout-2009} give an
unusual denotational model of SSA in terms of the iteration structure
of a program, which they use to better understand the loop-closing
$\phi$-nodes found both in the gated SSA representation as well as
practical compilers such as GCC.  \citet{garbuzov-structural-cfg-2018}
exhibit a correspondence between an operational semantics for SSA, and
an operational semantics for the call-by-push-value calculus of
\citet{cbpv}. Despite working over an operational semantics, they use
the normal-form bisimulations of \citet{lassen-bisim} to derive an
equational theory for use in justifying optimisations. Their
operational semantics does contain not contain any effects, which
means that it cannot justify effect-dependent optimisations such as
eliminating redundant writes.

Our work is the first to validate the equational theory of SSA against
a weak memory semantics. We focused on denotational models of weak
memory, since they make it easier to establish equational properties
than the operational and axiomatic models. However, for our proof to
go through, we needed a number of features of the semantics which are
sometimes difficult to establish in the weak memory world. First,
sequencing must be associative (i.e.,
$(t_1; t_2); t_3 \equiv t_1; (t_2; t_3)$).  Second, the semantics must
support loops, including the equation that a while-loop is equal to
its unrolling. Finally, sums need to be distributive (e.g., in the
boolean case, $(A \otimes \mb{2}) \simeq (A + A)$).

As a result, we focused on the TSO semantics of \citet{sparky}, which
does have all the properties we need. For the weaker ARM-style models
(which admit load buffering), there seem to be a number of models
which are "near misses", in that they have some, but not all, the
properties we need. The event structure model of
\citet{paviotti-modular-relaxed-dep-20} is formulated in terms of
step-indexing, so satisfies all of our needed properties except the
loop unrolling property -- step-indexing means it is a refinement
rather than an equation. The pomsets with preconditions model of
\citet{jagadeesan-pwp-20} does not exactly satisfy the associativity
of semicolon, and while the pomsets with predicate transformers model
of \citet{leaky-semicolon} does satisfy associativity, it does not yet
support loops.

We do not think it would be prohibitively difficult to extend these
models to support the required features, and we believe our results
suggest that it would be have a high payoff: those features would
suffice to build a model of our IR, giving us tight results
validating the transformations compilers want against the low-level
specification of the actual machine.
